---
title: "Integrative meta-analysis of epigenome-wide association studies identifies genomic and epigenomics differences in the brain and the blood in Alzheimer’s disease"
subtitle: 'ADNI dataset'
author:
  - Tiago Chedraoui Silva^[University of Miami]
  - Lily Wang^[University of Miami]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmarkdown::html_document:
    highlight: breezedark
    theme: lumen
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
knitr::opts_knit$set(root.dir = "./../../")
```

# Setup

```{R Setup, message = FALSE, results = 'hide'}
library(dplyr)
library(SummarizedExperiment)
library(ExperimentHub)
library(ADNIMERGE)
library(dplyr)
library(readxl)
library(wateRmelon)
library(minfi)
cohort <- "ADNI"
dir.base <- getwd()
dir.data <- file.path(dir.base,"datasets/",cohort,"/data/DNA_methylation") 
dir.data.clinical <- file.path(dir.base,"datasets/",cohort,"/data/Clinical") 
dir.data.raw <- file.path(dir.base,"datasets/",cohort,"/data/DNA_methylation/idat") 
dir.data.processed <- file.path(dir.base,"datasets/",cohort,"/data/DNA_methylation/processed") 
dir.data.aux <- file.path(dir.base,"datasets/Aux/") 
dir.data.pca <- file.path(dir.data,"/pca_filtering/") 
dir.results <- file.path(dir.base,"analysis_results",cohort)
dir.results.lm <- file.path(dir.results,"/Linear_mixed_model")
dir.results.glm <- file.path(dir.results,"/Logistic_regression_model")
for(p in grep("dir",ls(),value = T)) dir.create(get(p),recursive = TRUE,showWarnings = FALSE)
```

# Data retrieval

Source: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE153712

```{r Data_retrieval, eval = FALSE}
RGSet <- read.metharray.exp(
  base = dir.data.raw,
  recursive = TRUE,
  verbose = TRUE
)
saveRDS(RGSet, paste0(dir.data.processed,"/RGSet.RDS"))
```

#  Data Pre-processing 

```{R pre_processing, eval = FALSE}
message(object.size(RGSet))
bs <- data.frame(bisulfiteConversion = bscon(RGSet))
bsFilteredOut <- row.names(bs)[bs$bisulfiteConversion < 85]
nb.samples <- ncol(RGSet)
RGSet.filtered <- RGSet[,!colnames(RGSet) %in% bsFilteredOut]
nb.samples.bc.filtered <-  ncol(RGSet.filtered)

save(
  RGSet.filtered, 
  bs, 
  nb.samples, 
  nb.samples.bc.filtered, 
  file =  file.path(dir.data.processed,"/RGSet_filtered_bs_min_85.rda")
)
```


```{R load_RGSet_filtered_bs_min_85, eval = TRUE}
load(file.path(dir.data.processed,"/RGSet_filtered_bs_min_85.rda"))
```

```{R, eval = FALSE}
bs$barcodes <- rownames(bs)
meta <- readxl::read_xlsx(file.path(dir.data,"ADNI_DNA_Methylation_SampleAnnotation_20170530.xlsx"))
meta.with.bs <- dplyr::left_join(bs,meta)
meta.with.bs <- meta.with.bs[order(-meta.with.bs$bisulfiteConversion),]
meta.with.bs$RID_Phase_Edata <- paste0(meta.with.bs$RID,"_",meta.with.bs$Phase,"_",meta.with.bs$Edate)
meta.with.bs$is_replicate <- duplicated(meta.with.bs$RID_Phase_Edata)

readr::write_csv(
  meta.with.bs,
  file = file.path(dir.data.processed,"ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs.csv")
)

meta.with.bs <- meta.with.bs %>% dplyr::filter(is_replicate == FALSE & !is.na(Slide) & bisulfiteConversion >= 85)
samples.keep <- meta.with.bs %>% dplyr::pull(barcodes)
RGSet.filtered <- RGSet.filtered[,samples.keep]

save(
  RGSet.filtered, 
  meta.with.bs,
  file = file.path(dir.data.processed,"/RGSet_filtered_bs_min_85_no_replicates_no_missing_info.rda")
)
```

## Remove samples with age < 65
```{R, load_bs_filtered, include = FALSE}
load(file.path(dir.data.processed,"RGSet_filtered_bs_min_85_no_replicates_no_missing_info.rda"))
nb.probes.before.norm <- nrow(RGSet.filtered)
```

```{R}
meta.with.bs$ID <- paste0(     
  meta.with.bs$RID,
  meta.with.bs$Phase,
  meta.with.bs$Edate
)

dnam.rid <- meta.with.bs$ID[match(colnames(RGSet.filtered), meta.with.bs$barcodes)]
clinical <- readr::read_csv(file.path(dir.data.clinical,"ADNIMERGE_downloaded_3-1-2021.csv"))
clinical$ID <- paste0(clinical$RID,clinical$COLPROT,clinical$EXAMDATE)
clinical$ID2 <- paste0(clinical$RID,clinical$COLPROT,clinical$VISCODE)
# it seems some samples do not have clinical data ?
dnam.rid <- dnam.rid[dnam.rid %in% clinical$ID] 
clinical <- clinical[match(dnam.rid,clinical$ID),]

clinical$PlateNumber <- meta.with.bs$PlateNumber[
    match(clinical$ID,meta.with.bs$ID)
]
clinical$bisulfiteConversion <- meta.with.bs$bisulfiteConversion[
    match(clinical$ID,meta.with.bs$ID)
]
clinical$Array  <- meta.with.bs$Array[
    match(clinical$ID,meta.with.bs$ID)
]

clinical$barcodes  <- meta.with.bs$barcodes[
    match(clinical$ID,meta.with.bs$ID)
]

clinical$DateDrawn <- meta.with.bs$DateDrawn[
    match(clinical$ID,meta.with.bs$ID)
]

rownames(clinical) <- clinical$barcodes
clinical$PlateNumber <- factor(clinical$PlateNumber)

demo <- readr::read_csv(file.path(dir.data.clinical,"PTDEMOG_3-1-2021.csv"))
clinical$birth_month <- demo$PTDOBMM[match(clinical$RID, demo$RID)]
clinical$birth_year <- demo$PTDOBYY[match(clinical$RID, demo$RID)]

# we don't have the day so we set to one
library(lubridate)
clinical$age_at_visit <- interval(
    as.Date(paste0(clinical$birth_month,"/1/",clinical$birth_year), "%m/%d/%Y"),
    clinical$DateDrawn
) %>% time_length(unit = "years")
```

```{R}
samples.min.age.65 <- clinical$barcodes[clinical$age_at_visit >= 65]
clinical.filtered.min.age_at_visit.65 <- clinical[clinical$age_at_visit >= 65,]
RGSet.bs.filtered.min.age_at_visit.65 <- RGSet.filtered[,samples.min.age.65]
```

```{R, include = FALSE}
rm(RGSet.filtered)
gc()
```

```{R save_RGSet.bs.filtered.min.age.65, eval = FALSE}
save(
  RGSet.bs.filtered.min.age_at_visit.65, 
  clinical.filtered.min.age_at_visit.65,
  file = file.path(dir.data.processed,"RGSet_bs_filtered_min_age_at_visit_65.rda")
)
```

```{R age_hist, eval = TRUE, echo = FALSE}
library(ggplot2)
ggpubr::gghistogram(data = clinical,x = "age_at_visit",color = "white",fill = "blue", bins = 50) + 
  geom_vline(xintercept = 65, color = "red", size=0.4) +
    geom_text(color = "red",
        size    = 5,
        data    = plyr::count(clinical$age_at_visit < 65) %>% filter(x == T),
        mapping = aes(x = -Inf, y = Inf, label = freq),
        hjust   = -1.05,
        vjust   = 2
    ) +
    geom_text(color = "blue",
        size    = 5,
        data    = plyr::count(clinical$age_at_visit >= 65) %>% filter(x == T),
        mapping = aes(x = Inf, y = Inf, label = freq),
        hjust   = 2,
        vjust   = 2
    )
```

## Probes QC

1. keep only probes that start with "cg"
2. drop probes that are on X/Y
3. drop probes where SNP with MAF >= 0.01 was present in the last 5 bp of the probe.
4. Remove smoking related probes

```{R, load_bs_filtered_min_65, include = FALSE}
load(file.path(dir.data.processed,"RGSet_bs_filtered_min_age_at_visit_65.rda"))
nb.probes.before.norm <- nrow(RGSet.bs.filtered.min.age_at_visit.65)
```


```{R, eval = FALSE}
library(minfi)
library(wateRmelon)
library(RPMM)
library(DMRcate)
detP <- detectionP(RGSet.bs.filtered.min.age_at_visit.65, type = "m+u")
failed.01 <- detP > 0.01
passedProbes <- rownames(failed.01)[rowMeans(failed.01) == 0]
sum(rowMeans(failed.01) == 0)  
save(passedProbes, file = file.path(dir.data.processed,"detectionP_passed_probes.rda"))
```

```{R}
load(file.path(dir.data.processed,"detectionP_passed_probes.rda"))
```

```{R, eval = FALSE}
betaSet <- getBeta(RGSet.bs.filtered.min.age_at_visit.65)
rm(RGSet.bs.filtered.min.age_at_visit.65)
gc()
betaSet <- betaSet[rownames(betaSet) %in% passedProbes, ]
betaSet <- betaSet[grep("cg",rownames(betaSet)),]

# snapshotDate(): 2021-03-30
betaSet <- rmSNPandCH(
  object = betaSet,
  dist = 5,
  mafcut = 0.01,
  and = TRUE,
  rmcrosshyb = TRUE,
  rmXY = TRUE
)
```


## Remove smoking probes

```{R, eval = FALSE}
smoking.file <- "https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5267325/bin/NIHMS817273-supplement-001506_-_Supplemental_Tables.xlsx"
if(!file.exists(file.path(dir.data.aux,basename(smoking.file)))) {
  downloader::download(smoking.file,file.path(dir.data.aux,basename(smoking.file)))
}
smoking <- readxl::read_xlsx(
    file.path(dir.data.aux,basename(smoking.file)),
    sheet = "02",
    skip = 2
)
smoking.sig.probes <- smoking %>%
  dplyr::filter(`P-value` < 1*10^(-7)) %>%
  dplyr::pull("Probe ID")

betaSet <- betaSet[!rownames(betaSet) %in% smoking.sig.probes,]

save(
  betaSet, 
  clinical.filtered.min.age_at_visit.65,
  file = file.path(dir.data.processed,"RGSet_bs_filtered_min_age_at_visit_65_before_norm.rda")
)

```

## Samples QC

### Quantile normalization

```{R, eval = FALSE}
library(lumi)
betaQN <- lumiN(x.lumi = betaSet, method = "quantile")
saveRDS(betaQN, file.path(dir.data.processed,"betaQN_min_age_at_visit_65.RDS"))
```

```{R, include = FALSE}
betaQN <- readRDS(file.path(dir.data.processed,"betaQN_min_age_at_visit_65.RDS"))
```

```{R}
boxplot(betaQN[,1:100])
```

### BMIQ normalization
```{R, message = FALSE, results = "hide"}
library(wateRmelon)
library(RPMM)
```

```{R, eval = TRUE}
annotType <- IlluminaHumanMethylationEPICanno.ilm10b4.hg19::Manifest
annotType$designTypeNumeric <- ifelse(annotType$Type == "I",1,2)
library(sm)

betaQNCompleteCol1 <- betaQN[complete.cases(betaQN[,1]), ]
annotTypeCompleteCol1 <- annotType[row.names(betaQNCompleteCol1), ]

sm.density.compare(
  betaQNCompleteCol1[,1],
  annotTypeCompleteCol1$designTypeNumeric
)
rm(betaQNCompleteCol1)
rm(annotTypeCompleteCol1)
```

```{R, eval = FALSE}
type12 <- annotType$designTypeNumeric[match(rownames(betaQN),rownames(annotType))]
set.seed (946)
doParallel::registerDoParallel(cores = 1)
betaQN_BMIQ <- plyr::aaply(
  betaQN, 2,
  function(x){
    norm_ls <- BMIQ(x, design.v = type12, plots = FALSE)
    return (norm_ls$nbeta)
  },.progress = "time",.parallel = FALSE
) %>% t()

saveRDS(betaQN_BMIQ, file.path(dir.data.processed,"ADNI_QNBMIQ_min_age_at_visit_65.RDS"))
```


### Compute cell type proportions 
```{R, eval = FALSE, include = TRUE}
betaQN_BMIQ <- readRDS(file.path(dir.data.processed,"ADNI_QNBMIQ_min_age_at_visit_65.RDS"))

library(EpiDISH)
data(centDHSbloodDMC.m)
BloodFrac.m <- epidish(beta.m = betaQN_BMIQ, ref.m = centDHSbloodDMC.m, method = "RPC")$estF
```

### Summarize Experiment

```{R, eval = FALSE}
library(ADNIMERGE)
library(dplyr)
library(readxl)

ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs <- readr::read_csv(
  file.path(dir.data,"ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs.csv")
) 
ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$ID <- paste0(     
  ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$RID,
  ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$Phase,
  ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$Edate
)
dnam.rid <- ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$ID[match(
  colnames(betaQN_BMIQ), ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$barcodes
  )
]



# clinical <- ADNIMERGE::adnimerge
clinical <- readr::read_csv(file.path(dir.data.clinical,"ADNIMERGE_downloaded_3-1-2021.csv"))
# clinical$DIAGNOSIS <- clinical$DX
clinical$ID <- paste0(clinical$RID,clinical$COLPROT,clinical$EXAMDATE)
clinical$ID2 <- paste0(clinical$RID,clinical$COLPROT,clinical$VISCODE)
# it seems some samples do not have clinical data ?
dnam.rid <- dnam.rid[dnam.rid %in% clinical$ID] 
clinical <- clinical[match(dnam.rid,clinical$ID),]


clinical$PlateNumber <- ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$PlateNumber[
  match(clinical$ID,ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$ID)
]
clinical$bisulfiteConversion <- ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$bisulfiteConversion[
  match(clinical$ID,ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$ID)
]
clinical$Array  <- ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$Array[
  match(clinical$ID,ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$ID)
]

clinical$barcodes  <- ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$barcodes[
  match(clinical$ID,ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$ID)
]

clinical$DateDrawn <- ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$DateDrawn[
  match(clinical$ID,ADNI_DNA_Methylation_SampleAnnotation_20170530_with_bs$ID)
]

rownames(clinical) <- clinical$barcodes

clinical <- merge(
  clinical, 
  BloodFrac.m,
  by.x = "row.names",
  by.y = "row.names",
  sort = FALSE
)
clinical$PlateNumber <- factor(clinical$PlateNumber)
rownames(clinical) <- clinical$Row.names
clinical$Row.names <- NULL

demo <- readr::read_csv(file.path(dir.data.clinical,"PTDEMOG_3-1-2021.csv"))
clinical$birth_month <- demo$PTDOBMM[match(clinical$RID, demo$RID)]
clinical$birth_year <- demo$PTDOBYY[match(clinical$RID, demo$RID)]

# we don't have the day so we set to one
library(lubridate)
clinical$age_at_visit <- interval(
  as.Date(paste0(clinical$birth_month,"/1/",clinical$birth_year), "%m/%d/%Y"),
  clinical$DateDrawn
) %>% time_length(unit = "years")


clinical <-  clinical %>% dplyr::select(
  c(barcodes, 
    RID, 
    COLPROT,
    VISCODE, 
    PTGENDER, 
    PTEDUCAT,
    DateDrawn  ,
    DX, 
    PlateNumber,
    birth_year,
    birth_month,
    age_at_visit, 
    bisulfiteConversion,
    B, 
    NK,   
    CD4T, 
    CD8T,   
    Mono,   
    Neutro, 
    Eosino,
    APOE4
  ))

library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(S4Vectors)
epic.hg19 <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
epic.hg19.gr <- epic.hg19 %>% makeGRangesFromDataFrame(
  start.field = "pos", end.field = "pos",keep.extra.columns = T
)

betaQN_BMIQ <- betaQN_BMIQ[rownames(betaQN_BMIQ) %in% names(epic.hg19.gr),]
rowData <- epic.hg19.gr[rownames(betaQN_BMIQ),]
data <- betaQN_BMIQ[,colnames(betaQN_BMIQ) %in% rownames(clinical)]
data <- data[,rownames(clinical)]

adni.se <- SummarizedExperiment::SummarizedExperiment(
  assays = SimpleList("DNAm" = data),
  rowData = rowData,
  metadata = list("genome" = "hg19"),
  colData = clinical
)

saveRDS(adni.se, file = file.path(dir.data.processed,"adni_se_min_age_at_visit_65.RDS"))
readr::write_csv(
  clinical %>% as.data.frame, 
  file = file.path(dir.data.processed,"ADNI_Se_colData.csv")
)
```


# Outliers detection - PCA analysis

Description: 

1. Estimate standard deviation for each probe
2. Select most variable probes (e.g. n = 50,000)
3. PCA plot
4. Filter outliers

```{R load_plot_pca}
adni.se <- readRDS(file.path(dir.data.processed,"adni_se_min_age_at_visit_65.RDS"))
### transform to m values
pheno_df <- colData(adni.se) %>% as.data.frame
pheno_df$sample <- colnames(adni.se)
nb.samples.before.pca <- ncol(adni.se)
# plotPCA and OrderDataBySd functions
devtools::source_gist("https://gist.github.com/tiagochst/d3a7b1639acf603916c315d23b1efb3e")
##### 1.Order matrix by most variable probes on top ############################
betaOrd_mat <- OrderDataBySd(SummarizedExperiment::assay(adni.se)) 
```



```{R, eval = FALSE}
pca <- prcomp(
  t(betaOrd_mat[1:50000,]),
  center = TRUE,
  scale = TRUE
)
d <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2])
meanPC1 <- mean (d$PC1)
sdPC1   <- sd (d$PC1)
meanPC2 <- mean (d$PC2)
sdPC2   <- sd (d$PC2)
out3sdPC1_1 <- meanPC1 - 3*sdPC1
out3sdPC1_2 <- meanPC1 + 3*sdPC1
out3sdPC2_1 <- meanPC2 - 3*sdPC2
out3sdPC2_2 <- meanPC2 + 3*sdPC2
d$outlier_PC1[d$PC1 >= out3sdPC1_1 & d$PC1 <= out3sdPC1_2] <- 0
d$outlier_PC1[d$PC1 < out3sdPC1_1 | d$PC1 > out3sdPC1_2] <- 1
d$outlier_PC2[d$PC2 >= out3sdPC2_1 & d$PC2 <= out3sdPC2_2] <- 0
d$outlier_PC2[d$PC2 < out3sdPC2_1 | d$PC2 > out3sdPC2_2] <- 1
write.csv(d, paste0(dir.data.pca, "ADNI_PCs_usingBetas_min_age_at_visit_65.csv"))
```

## PCA plot

```{R pca_plot, eval = TRUE}
library(ggplot2)
library(ggrepel)
### beta values
byStatus <- plotPCA(
  dataset = "ADNI: beta values",
  expSorted_mat = betaOrd_mat,
  pheno = pheno_df,
  group_char = "DX",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)
bySex <- plotPCA(
  dataset = "ADNI: beta values",
  expSorted_mat = betaOrd_mat,
  pheno = pheno_df,
  group_char = "PTGENDER",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)
pheno_df$PlateNumber <- as.factor(pheno_df$PlateNumber)
byPLATE <- plotPCA(
  dataset = "ADNI: beta values",
  expSorted_mat = betaOrd_mat,
  pheno = pheno_df,
  group_char = "PlateNumber",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)
```

## Filter samples by PCA 
```{R save_pca, eval = FALSE}
noOutliers <- d[which(d$outlier_PC1 == 0 & d$outlier_PC2 == 0), ]
adni.se <- adni.se[, rownames(noOutliers)]
saveRDS(adni.se, file.path(dir.data.pca, "ADNI_QNBMIQ_PCfiltered_min_age_at_visit_65.RDS"))
writexl::write_xlsx(
  colData(adni.se) %>% as.data.frame(),
  path =  file.path(dir.data.pca, "ADNI_QNBMIQ_PCfiltered_min_age_at_visit_65_metadata.xlsx")
)
```


```{R, include = FALSE}
rm(betaOrd_mat)
gc()
```

# Summary after QC steps

## Data and metadata
```{R}
adni.se <- readRDS(file.path(dir.data.pca, "ADNI_QNBMIQ_PCfiltered_min_age_at_visit_65.RDS"))
nb.samples.after.pca <- ncol(adni.se)
nb.probes.after.norm <- nrow(adni.se)
```


## Numbers of samples and probes removed in each step


```{R, eval = TRUE, echo=FALSE}
df.samples <- data.frame(
  "Number of samples" =  c(
    nb.samples, 
    nb.samples.bc.filtered, 
    nb.samples.before.pca, 
    nb.samples.after.pca
  ),
  "Description" = c(
    "Number of samples before bisulfiteConversion filter",
    "Number of samples after bisulfiteConversion filter",
    "Number of samples before PCA",
    "Number of samples after PCA"
  )
)    
df.samples 
```

```{R summary_table, eval = TRUE, echo=FALSE}
# Create summary table
df.probes <- data.frame(
  "Number of probes" = c(
    nb.probes.before.norm,
    length(passedProbes),
    nb.probes.after.norm
  ),
  "Description" = c(
    "total number of probes in raw data",
    "probes with detection P < 0.01 for all samples",
    "only probes that start with cg, no crosshyb, no X, no Y, no smoking, mafcut = 0.01"
  )
)
df.probes %>% gt::gt() 
```

# Checking the data

```{R}
se <- readRDS(file.path(dir.data.pca, "ADNI_QNBMIQ_PCfiltered_min_age_at_visit_65.RDS"))
```

## show probes not starting with cg are removed
```{R check_cg}
grep("cg",rownames(se),invert = T)
```

## show probes on X/Y are removed
```{R check_xy}
c("chrX", "chrY") %in% (seqnames(se) %>% as.character())
```

## show probes with SNP with MAF >= 0.01 and in the last 5bp were removed (compared with DMRcateData package probes)
```{R}
eh <- ExperimentHub()
snpsall <- eh[["EH3130"]]
any(snpsall$probe[snpsall$mafs > 0.01 & snpsall$distances <= 5] %>% as.character() %in% rownames(se)) 
crosshyb <- eh[["EH3129"]]
any(crosshyb %in% rownames(se))
XY.probes <- eh[["EH3131"]]
any(XY.probes %in% rownames(se)) 
```

## select a few smoking probes, and show they are removed
```{R}
any(
  c("cg16145216", "cg19406367",
    "cg05603985", "cg10768063",
    "cg14099685", "cg12513616",
    "cg04085571", "cg23648810"
  ) %in% rownames(se)
)
```

## show quantiles for each sample is the same
```{R quantile_box_plot}
boxplot(assay(se[,1:10]))
plyr::adply(assay(se[,1:10]),.margins = 2,.fun = function(x){quantile(x)})
```


## check cell type proportions adds up to 1
```{R}
table(rowSums(colData(se)[,c("B","NK","CD4T","CD8T","Mono","Neutro","Eosino")] %>% as.matrix()))
```

## show plate is a factor
```{R}
class(se$PlateNumber)
```

## check outliers samples are not in datset anymore
```{R}
outliers <- readr::read_csv(paste0(dir.data.pca, "ADNI_PCs_usingBetas_min_age_at_visit_65.csv"),col_types = readr::cols()) %>%
  dplyr::filter(outlier_PC1 == 1 | outlier_PC2 == 1) %>% pull(X1) 
head(outliers)
head(colnames(se))
any(outliers %in% colnames(se))
```

## Low varying probes were removed 

```{R}
any(rowSums(assay(se) > 0.85) == ncol(se))
any(rowSums(assay(se) < 0.15) == ncol(se))
```

## for one sample, check distribution of type I and type II probes
```{R checking_part2, eval = TRUE}
library(sm)

annotType <- IlluminaHumanMethylationEPICanno.ilm10b4.hg19::Manifest
annotType$designTypeNumeric <- ifelse(annotType$Type == "I",1,2)
probes <- assay(se)[,1] %>% na.omit()
sm.density.compare(
  probes,
  annotType[names(probes),]$designTypeNumeric
)

probes <- assay(se)[,2] %>% na.omit()
sm.density.compare(
  probes,
  annotType[names(probes),]$designTypeNumeric
)

```


```{R checking_part5, include = FALSE}
rm(se)
```


# Analysis
```{R load_analysis_data, eval = FALSE}
adni.se <- readRDS(file.path(dir.data.pca, "ADNI_QNBMIQ_PCfiltered_min_age_at_visit_65.RDS"))
```


```{R sas, eval = FALSE}
str_data <- function(adni.se.selected){
  
  mvalues <- log2(assay(adni.se.selected)/(1 - assay(adni.se.selected)))
  mvalues <- mvalues %>% t %>% as.data.frame()
  mvalues$barcodes <- rownames(mvalues)
  mvalues.pivot <- pivot_longer(mvalues,!barcodes,names_to = "cpg", values_to = "mvalues")
  
  beta <- assay(adni.se.selected)
  beta <- beta %>% t %>% as.data.frame()
  beta$barcodes <- rownames(beta)
  beta.pivot <- pivot_longer(beta,!barcodes,names_to = "cpg", values_to = "beta")
  
  
  #  0 = control, 1 = AD, 2 = MCI
  pheno <- data.frame(colData (adni.se.selected))
  pheno$DIAGNOSIS <- factor(pheno$DX, levels = c("CN", "MCI", "Dementia"))
  pheno <- pheno[!is.na(pheno$DIAGNOSIS),]
  pheno$DIAGNOSIS_merged <- ifelse(pheno$DIAGNOSIS == "CN", 0, 1)
  pheno$DIAGNOSIS_merged <- factor(pheno$DIAGNOSIS_merged, levels = c(0, 1))
  pheno$barcodes <- row.names(pheno)
  
  # add indicator for first and last visit
  pheno$visit <- sub("m",  "", pheno$VISCODE)
  pheno$PlateNumber <- factor(pheno$PlateNumber)
  
  pheno <- pheno %>% dplyr::select(
    c(barcodes, RID, VISCODE, PTGENDER, age_at_visit, visit, 
      DIAGNOSIS, DIAGNOSIS_merged, PlateNumber,
      B, NK,   CD4T, CD8T,   Mono,   Neutro, Eosino
    )
  )
  final <- dplyr::inner_join(pheno,beta.pivot) %>%  dplyr::inner_join(mvalues.pivot)
  final <- final[order(final$cpg,final$RID,final$VISCODE),]
  final
}

for(start in seq(1,nrow(adni.se),by = 1000)){
  print(start)
  end <- start + 1000 - 1
  idx <- which(start == seq(1,nrow(adni.se),by = 1000))
  if(end > nrow(adni.se)) end <- nrow(adni.se)
  adni.se.selected <- adni.se[start:end,adni.se$DX %in% c("Dementia","CN")]
  final <- str_data(adni.se.selected)
  final$DIAGNOSIS_merged <- final$VISCODE <- final$visit <- NULL
  readr::write_csv(
    final,
    file = paste0(
      dir.lw,
      "AD_CN/DATA_for_SAS_models//",
      "ADNI_1000_cpgs_",idx,".csv"
      )
  )
}

```

## Logistic regression model


### AD vs CN
```{R, eval = FALSE}
files <- dir(
  file.path(dir.base,"LW/ADNI_GLMM/AD_CN/results/"),
  full.names = T,
  recursive = T,
  pattern = "_[0-9]*.csv"
)
length(files)
glmm <- plyr::adply(
  .data = files,
  .margins = 1,
  .fun = function(f) {
    readr::read_csv(f,col_types = readr::cols())
  },.progress = "time",.id = NULL) %>% unique()  
readr::write_csv(
  glmm, 
  file = file.path(dir.results.glm,"ADNI_AD_vs_CN_glmm_using_beta_SAS.cvs")
)
```

## Annotate results
```{R annotate_functions, eval = FALSE}
message("o Stating to annotate")
adni.se <- readRDS(file.path(dir.data.pca, "ADNI_QNBMIQ_PCfiltered_min_age_at_visit_65.RDS"))
load(file.path(dir.data.aux,"great_EPIC_array_annotation.rda"))
load(file.path(dir.data.aux,"E073_15_coreMarks_segments.rda"))
load(file.path(dir.data.aux,"meta_analysis_cpgs.rda"))

annotate_results <- function(result, data.se){
  #result <- result[result$cpg %in% rownames(data.se),]
  result$fdr <- p.adjust(result$Probt,method = "fdr")
  result <- cbind(
    result,
    rowRanges(data.se)[result$cpg] %>% as.data.frame() %>% dplyr::select(1:4)
  )
  result$Islands.UCSC.Relation_to_Island <- IlluminaHumanMethylationEPICanno.ilm10b4.hg19::Islands.UCSC[result$cpg,"Relation_to_Island"]
  result$UCSC_RefGene_Name <- IlluminaHumanMethylationEPICanno.ilm10b4.hg19::Other[result$cpg,"UCSC_RefGene_Name"]       
  result$UCSC_RefGene_Group <- IlluminaHumanMethylationEPICanno.ilm10b4.hg19::Other[result$cpg,"UCSC_RefGene_Group"]     
  
  result <- dplyr::left_join(result, great,by = c("seqnames","start","end","cpg"))
  
  hits <- findOverlaps(rowRanges(data.se)[result$cpg],ChmmModels.gr) %>% as.data.frame()
  hits$state <- ChmmModels.gr$state[hits$subjectHits]
  hits$cpg <- result$cpg[hits$queryHits]
  result <- merge(result, hits[,c("state","cpg")], all.x = TRUE, sort = FALSE, by = "cpg")
  
  result$sig.in.brain <- result$cpg %in% brain.meta.analysis.cpgs
  return(result)
}
```

```{R annotate, eval = FALSE}
logistic_mixed_model_ad_cn <- readr::read_csv( 
  file.path(dir.results.glm,"ADNI_AD_vs_CN_glmm_using_beta_SAS.cvs")
)
logistic_mixed_model_ad_cn.annotated <- logistic_mixed_model_ad_cn %>% annotate_results(adni.se)
readr::write_csv(
  x = logistic_mixed_model_ad_cn.annotated,
  file = file.path(dir.results.glm,"ADNI_AD_vs_CN_glmm_using_beta_all_annotated.cvs"),
)
```

```{R read_results, include = FALSE}
logistic_mixed_model_ad_cn.annotated <- readr::read_csv(
  file.path(dir.results.glm,"ADNI_AD_vs_CN_glmm_using_beta_all_annotated.cvs")
)
```

# Calculate inflation factors and Bacon correction for final results

## Auxiliary functions

```{R inflation}
library(dplyr)
library(bacon)
library(GWASTools)
estimation_of_inflation <- function(data){
  ### 1. Compute genomic inflation factor before bacon adjustment
  data$zvalue <- data$Estimate / data$StdErr
  #data$zvalue <- qnorm(pt(data$tvalue, df))
  data$chisq <- (data$zvalue) ^ 2
  # inflation factor - last term is median from chisq distrn with 1 df  
  inflationFactor <- median(data$chisq,na.rm = TRUE) / qchisq(0.5, 1)
  print("lambda")
  print(inflationFactor)
  # genome-wide sig cpgs
  #sig <- ifelse(data$`Pr(>|t|)` < 2.4e-7, 1, 0)
  # table(sig)  # 1 sig
  ### 2. bacon analysis
  bc <- bacon(
    teststatistics = NULL,
    effectsizes =  data$Estimate,
    standarderrors = data$StdErr,
    na.exclude = TRUE
  )
  # inflation factor
  print("lambda.bacon")
  print(inflation(bc))
  
  ### 3. Create final dataset
  data.with.inflation <- data.frame(
    data,
    Estimate.bacon = bacon::es(bc),
    StdErr.bacon = bacon::se(bc),
    pValue.bacon = pval(bc),
    fdr.bacon = p.adjust(pval(bc), method = "fdr"),
    stringsAsFactors = FALSE
  )
  data.with.inflation <- data.with.inflation %>% dplyr::select(-c(zvalue, chisq))
  return(
    list(
      "data.with.inflation" = data.with.inflation,
      "inflationFactor" =   inflationFactor,
      "estimatedInflation" = inflation(bc)
    )
  )
}

plot_inflation <- function(data, main){
  qqPlot(
    data$data.with.inflation$Probt, ci = TRUE, 
    main = paste0(main)
  ) + text(
    x = 0.5, 
    y = -log10(min(data$data.with.inflation$Probt)) - 0.5,
    labels = bquote(lambda == .(format(data$inflationFactor,digits = 4))),
    pos = 4
  ) + text(
    x = 0.5, 
    y = -log10(min(data$data.with.inflation$Probt)) - 1,
    labels = bquote(lambda[bacon] == .(format(data$estimatedInflation,digits = 4))),
    pos = 4
  )
}
```

```{R}
logistic_mixed_model_ad_cn.with.inflation <- estimation_of_inflation(data = logistic_mixed_model_ad_cn.annotated)
dim(logistic_mixed_model_ad_cn.with.inflation$data.with.inflation)
```

```{R, eval = F}
readr::write_csv(
  x = logistic_mixed_model_ad_cn.with.inflation$data.with.inflation,
  file = file.path(dir.results.glm,"ADNI_AD_vs_CN_glmm_using_beta_with_inflation.csv")
)
```

```{R}
plot_inflation(logistic_mixed_model_ad_cn.with.inflation, "ADNI AD vs CN - single cpg -  Logistic regression model")

qqPlot(
  logistic_mixed_model_ad_cn.with.inflation$data.with.inflation$pValue.bacon, ci = TRUE, 
  main = paste0("AIBL AD vs CN - single cpg - logistic regression mode - pValue.bacon")
) 

```

# Session information
```{R}
devtools::session_info()
```