---
title: "Figure 2 numbers"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
knitr::opts_knit$set(root.dir = "./../../")
library(dplyr)
```

# Cpg

## Row 1

### Blood samples meta-analysis  (at P< 0.05, with bacon-correction)
```{r}
blood.meta_analysis <- readr::read_csv(
  "analysis_results/meta_analysis/Logistic_regression_model/AD_vs_CN/meta_analysis_glm_fixed_effect_ADNI_and_AIBL_AD_vs_CN_single_cpg_annotated.csv",
  col_types = readr::cols()
)

blood.meta_analysis.sig <- blood.meta_analysis %>% dplyr::filter(pVal.final.bacon < 0.05)   
nrow(blood.meta_analysis.sig)
```


### Brain samples meta-analysis (at 5% FDR, with bacon-correction)

```{R}
brain.meta_analysis <- readr::read_csv(
  "datasets/brain_meta_analysis//meta_analysis_single_cpg_sig_no_crossHyb_smoking_df_bacon.csv",
  col_types = readr::cols()
)

brain.meta_analysis.sig <- brain.meta_analysis %>% dplyr::filter(fdr.bacon < 0.05)   
nrow(brain.meta_analysis.sig)
```

### Cross-tissue meta-analysis (at 5% FDR, with bacon-correction)

```{R}
cross.tissue <- readr::read_csv(
  file = "analysis_results/cross_meta_analysis/cross_tissue_meta_analysis_glm_using_AD_vs_CN_single_cpg.csv",
  col_types = readr::cols()
)
cross.tissue.sig <- cross.tissue %>% dplyr::filter(fdr < 0.05)   
nrow(cross.tissue.sig)
```

## Row 2

### Overlap of brain & cross-tissue meta-analyses 
```{r}
intersect(brain.meta_analysis.sig$cpg,cross.tissue.sig$cpg) %>% length
```

### Overlap of blood & cross-tissue meta-analyses 
```{r}
intersect(blood.meta_analysis.sig$cpg,cross.tissue.sig$cpg) %>% length
```

## Row 3

### Overlap of brain, blood and cross-tissue meta-analyses

```{R} 
intersect(blood.meta_analysis.sig$cpg,cross.tissue.sig$cpg) %>% 
  intersect(brain.meta_analysis.sig$cpg) %>% length
```

# DMR

## Row 1

### Blood samples meta-analysis  (at P< 0.05, with bacon-correction)
```{R, eval = TRUE}
blood.meta_analysis.dmr <- readr::read_csv(
  "analysis_results/meta_analysis/DMR_Logistic_regression_model/AD_vs_CN/meta_analysis_glm_fixed_effect_ADNI_and_AIBL_AD_vs_CN_DMR_annotated.csv",
  col_types = readr::cols()
  )


blood.meta_analysis.dmr.sig <- blood.meta_analysis.dmr %>% dplyr::filter(pVal.final.bacon < 0.05)
nrow(blood.meta_analysis.dmr.sig)
```

### Brain samples meta-analysis (at 5% FDR, with bacon-correction)

```{R}
brain.meta_analysis.dmr <- readr::read_csv(
  "code/others/brain_DMR_meta_analysis_with_bacon_inflation.csv",
  col_types = readr::cols()
) %>% dplyr::mutate(fdr.bacon = p.adjust(pVal.final.bacon,method = "fdr"))

brain.meta_analysis.dmr.sig <- brain.meta_analysis.dmr %>%  dplyr::filter(fdr.bacon < 0.05)
nrow(brain.meta_analysis.dmr.sig)
```
### Cross-tissue meta-analysis (at 5% FDR, with bacon-correction)

```{R}
cross.tissue.dmr <- readr::read_csv(
  file = paste0("analysis_results/cross_meta_analysis/cross_tissue_meta_analysis_glm_using_AD_vs_CN_DMR.csv"),
  col_types = readr::cols()
)
cross.tissue.dmr.sig <- cross.tissue.dmr %>% dplyr::filter(fdr < 0.05)   
nrow(cross.tissue.dmr.sig)
```

## Row 2

### Overlap of brain & cross-tissue meta-analyses 

```{r}
intersect(brain.meta_analysis.dmr.sig$inputRegion,cross.tissue.dmr.sig$inputRegion_chr) %>% length
```

### Overlap of blood & cross-tissue meta-analyses 

```{r}
intersect(blood.meta_analysis.dmr.sig$cpg,cross.tissue.dmr.sig$inputRegion_chr) %>% length
```

## Row 3

### Overlap of brain, blood and cross-tissue meta-analyses

```{R} 
intersect(brain.meta_analysis.dmr.sig$inputRegion,cross.tissue.dmr.sig$inputRegion_chr) %>% 
  intersect(blood.meta_analysis.dmr.sig$cpg) %>% length
```

