---
title: "Evaluating effect of APOE "
date: "`r Sys.Date()`"
output: 
 html_document:
    theme:
      bg: "#202123"
      fg: "#B8BCC2"
      primary: "#EA80FC"
      base_font: !expr bslib::font_google("Barlow")
    highlight: zenburn
    toc: true
    df_print: paged
    code_download: false
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
vignette: >
    %\VignetteIndexEntry{Evaluating effect of APOE}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
knitr::opts_knit$set(root.dir = "../../")
```

# Libs and Paths
```{r libs, include=T,message=F,warning=F}
library(readr)
library(dplyr)
library(fgsea)
library(msigdbr)
library(SummarizedExperiment)
library(tidyverse)
library(caret)
library(factoextra)
library(pROC)
library(minfi)
```

# Get Cpgs

## AD associated cpgs (AD vs CN) - CpGs

```{r}
#---------------------------------------------------------------------------------------------------------------
# (a) AD-associated CpGs: include both 50 AD-associated CpGs + all CpGs from the 9 DMRs in Table 2 
#---------------------------------------------------------------------------------------------------------------
AD_vs_CN <- readxl::read_xlsx(
  "DRAFT-TABLES_FIGURES_4-17-2021/_Supp Table 2 final_AD_vs_CN-selcted-columns-formatted-V3.xlsx",skip = 3
)

cpgs.ad.cn <- AD_vs_CN$cpg
length(cpgs.ad.cn)
```

## AD associated cpgs (AD vs CN) - DMR
```{r}

dmr.ad.cn <- c(
  "chr1:205819345-205819464",
  "chr20:36148672-36148861",
  "chr1:206786170-206786181",
  "chr20:36149081-36149232",
  "chr13:21578684-21578734",
  "chr14:56777451-56777526",
  "chr7:2728841-2728913",
  "chr1:202172848-202172913",
  "chr1:223566643-223566710"
)
length(dmr.ad.cn)
dmr.ad.cn.cpgs.hm450 <- lapply(dmr.ad.cn, function(x) {
  print(x);
  cpgs <- tryCatch({coMethDMR::GetCpGsInRegion(x,genome = "hg19",arrayType = "450k")}, error = function(e){return(NULL)})
  cpgs
}) %>% unlist %>% unique

dmr.ad.cn.cpgs.epic <- lapply(dmr.ad.cn, function(x) {
  print(x);
  cpgs <- tryCatch({coMethDMR::GetCpGsInRegion(x,genome = "hg19",arrayType = "EPIC")}, error = function(e){return(NULL)})
  cpgs
}) %>% unlist %>% unique


ad.associated.cpgs.with.dmr.cpgs <- c(cpgs.ad.cn,dmr.ad.cn.cpgs.epic,dmr.ad.cn.cpgs.hm450) %>% unique
length(ad.associated.cpgs.with.dmr.cpgs)
```


## AD prioritized Cpgs
```{R}
#---------------------------------------------------------------------------------------------------------------
# (b) cross-tissue (i.e., prioritized CpGs): include both 97 CpGs + all CpGs from the 10 DMRs in Fig 2
#---------------------------------------------------------------------------------------------------------------
cpgs.prioritized <- readxl::read_xlsx(
  "DRAFT-TABLES_FIGURES_4-17-2021/_Supp Table 3 prioritized-CpGs-crossTissue_brain_blood-clean.xlsx",skip = 3
)
cpgs.prioritized  <- cpgs.prioritized$CpG %>% na.omit() %>% as.character
length(cpgs.prioritized)
```


## AD prioritized DMRs
```{R}

dmr.prioritized <- c(
  "chr16:57405979-57406511",
  "chr6:31554829-31555016",
  "chr6:30853948-30854233",
  "chr15:39871808-39872186",
  "chr6:166876490-166877038",
  "chr1:167090618-167090757",
  "chr17:7832680-7832943",
  "chr17:62009607-62009835",
  "chr12:125028166-125028339",
  "chr10:682693-682871"
)
length(dmr.prioritized)

dmr.prioritized.cpgs.hm450 <- lapply(dmr.prioritized, function(x) {
  print(x);
  cpgs <- tryCatch({coMethDMR::GetCpGsInRegion(x,genome = "hg19",arrayType = "450k")}, error = function(e){return(NULL)})
  cpgs
}) %>% unlist %>% unique

dmr.prioritized.cpgs.epic <- lapply(dmr.prioritized, function(x) {
  print(x);
  cpgs <- tryCatch({coMethDMR::GetCpGsInRegion(x,genome = "hg19",arrayType = "EPIC")}, error = function(e){return(NULL)})
  cpgs
}) %>% unlist %>% unique

ad.prioritized.cpgs.with.dmr.cpgs <- c(cpgs.prioritized,dmr.prioritized.cpgs.epic,dmr.prioritized.cpgs.hm450) %>% unique
length(ad.prioritized.cpgs.with.dmr.cpgs)
```


# Data

```{R data}
#-----------------------------------------------------------------------------
# ADNI
#-----------------------------------------------------------------------------
cohort <- "ADNI"
dir.base <- getwd()
dir.data <- file.path(dir.base,"datasets/",cohort,"/") 
dir.data.pca <- file.path(dir.data,"/data/DNA_methylation/pca_filtering/") 
adni.se <- readRDS(
  file.path(dir.data.pca, "ADNI_QNBMIQ_PCfiltered_min_age_at_visit_65.RDS")
)

adni.se <- adni.se[,!adni.se$DX %in% "MCI" ]
adni.se <- adni.se[,!is.na(adni.se$DX)]
table(adni.se$DX)
# Alzheimer's disease     healthy control 
#  135                    356 

file.age.adni <- file.path(dir.data,"/data/DNA_methylation/ADNI_Predicted_age.xlsx")
if(!file.exists(file.age.adni)){
  age <- age.predictor(assay(adni.se))
  tab.age <- age$age.pred
  tab.age$Sample <- rownames(tab.age)
  writexl::write_xlsx(tab.age,file.age.adni)
} else {
  tab.age <- readxl::read_xlsx(file.age.adni)
}
tab.age$barcodes <- tab.age$Sample


pheno <- colData(adni.se) %>% merge(tab.age,sort = FALSE)
pheno$DIAGNOSIS <- factor(
  ifelse(pheno$DX == "CN", "healthy control", "Alzheimer's disease"),
  levels = c("healthy control", "mild cognitive impairment", "Alzheimer's disease")
) %>% droplevels()
pheno$SEX <- pheno$PTGENDER  %>% factor()
pheno$AGE <- pheno$age_at_visit
pheno$age.pred.Elastic_Net <- pheno$Elastic_Net
pheno$DIAGNOSIS <- droplevels(pheno$DIAGNOSIS)
plyr::count(pheno$DIAGNOSIS)

# Keep only last visit 
pheno <- pheno %>% as.data.frame %>%  dplyr::group_by(RID) %>% filter(age_at_visit == max(age_at_visit)) 
adni.se <- adni.se[,adni.se$barcodes %in% pheno$barcodes]

# Put them in the same order
adni.se <- adni.se[,pheno$barcodes]
all(colnames(adni.se) == pheno$barcodes)

plyr::count(pheno$DIAGNOSIS)
```

# MRS

```{R mrs}
#---------------------------------------------------------------------------------------------------------------
# 2. Polygenic methylation risk scores (MRS score)
# (1) compute these as sum of beta values weighted by meta-analysis effect estimate (column “estimate.bacon”) in meta-analysis results file
#---------------------------------------------------------------------------------------------------------------
get_MRS <- function(cpgs){
  meta.analysis.results <- readr::read_csv("analysis_results/meta_analysis/Logistic_regression_model/AD_vs_CN/meta_analysis_glm_fixed_effect_ADNI_and_AIBL_AD_vs_CN_single_cpg_annotated.csv")
  MRS.weights <- meta.analysis.results$estimate.bacon[match(c(cpgs),meta.analysis.results$cpg)]
  names(MRS.weights) <- c(cpgs)
  MRS.weights %>% hist %>% plot
  
  # Remove if Cpgs were not evaluated  in meta-analysis
  MRS.weights <- MRS.weights[!is.na(MRS.weights)] 
  
  # Only keep the ones in ADNI
  common.cpgs <- intersect(rownames(adni.se),names(MRS.weights))
  
  # Calculate the MRS for each sample beta dot product weights
  # MRS.weights
  MRS <- MRS.weights[common.cpgs] %*% assay(adni.se)[common.cpgs,]
  
  print(all(colnames(MRS) == pheno$barcodes))
  return(MRS)
}

pheno <- cbind(pheno,
               data.frame(
                 "MRS_prioritized_cpgs" = get_MRS(cpgs.prioritized) %>% as.numeric(),
                 "MRS_associated_cpgs" = get_MRS(cpgs.ad.cn) %>% as.numeric() 
               )
)

pheno$MRS_prioritized_cpgs %>% hist
pheno$MRS_associated_cpgs %>% hist
```

# Evaluate AUC

##  Aux function to get AUC using the EIC package + glm fitted model

```{R get_auc}
get_auc <- function(model.formula = model.formula,data.training = NULL ,data.testing = NULL){
  model <- glm(
    formula = model.formula,
    data = data.training, 
    family = binomial
  )
  #-----------------------------------------------------------------------------
  #  Testing model
  #-----------------------------------------------------------------------------
  data.testing$probabilities <- model %>% predict(data.testing, type = "response")
  result <- roc(DIAGNOSIS ~ probabilities, data = data.testing,print.auc=TRUE)
  auc <- result$auc
  
  return(auc)
}
```

## Set models
```{R}
#-----------------------------------------------------------------------------
# Models
#-----------------------------------------------------------------------------
set.seed(12345)
var.predictors.no.apoe <- c("MRS", "AGE", "SEX", "B", "NK" ,"CD4T", "CD8T","Mono", "Neutro")
var.predictors.with.apoe <- c(var.predictors.no.apoe, "APOE4")
var.response <- "DIAGNOSIS"
model1 <- paste0(var.response, " ~ ",  paste(var.predictors.no.apoe,collapse = " + "))
model2 <- paste0(var.response, " ~ ",  paste(var.predictors.with.apoe,collapse = " + "))

data <- pheno[,c("DIAGNOSIS","APOE4","MRS_associated_cpgs","MRS_prioritized_cpgs", "AGE", "SEX", "B", "NK" ,"CD4T", "CD8T","Mono", "Neutro")]
```

## Split data
```{R}
#-----------------------------------------------------------------------------
# Split data in 10 balanced groups
#-----------------------------------------------------------------------------
list.of.partions <- createFolds(data$DIAGNOSIS,k = 10)
df.foldid <- list.of.partions %>% plyr::ldply(.fun = function(x) data.frame(x)) %>% dplyr::arrange(x)
print(table(df.foldid$.id,data$DIAGNOSIS[df.foldid$x]))
```

## Fit model and test

For each fold We will make the following (10x for each partition):
-  train on 9 partitions   
-  test on the one left outside
-  get AUC
Then we will get the average AUC.

```{R tab}
df <- list("Model1" = model1,"Model2" = model2) %>% purrr::map_dfr(.f = function(model) {
  print(model)
  list("MRS_associated_cpgs" = "MRS_associated_cpgs","MRS_prioritized_cpgs" = "MRS_prioritized_cpgs") %>% purrr::map_dfr(.f = function(col.mrs ) {
    
    print(col.mrs)
    unique(df.foldid$.id) %>% sort %>% purrr::map_dfr(.f = function(fold){
      print(fold)
      data$MRS <- data[,col.mrs, drop = T]
      data.testing <- data[df.foldid$x[df.foldid$.id == fold],]
      data.training <- data[df.foldid$x[df.foldid$.id != fold],]
      auc <- get_auc(model.formula = model,data.training = data.training, data.testing = data.testing)
      data.frame("MRS" = col.mrs,"Model" = model,"Fold_ID" = fold, "AUC" = auc %>% as.numeric)
    })
  })
})


summary.tab <- df %>% dplyr::group_by(MRS,Model) %>% summarise(max(AUC), min(AUC),mean(AUC),sd(AUC))
writexl::write_xlsx(summary.tab,path = "Eval_effect_APOE_summary_tab.xlsx")
summary.tab %>% gt::gt()
summary.tab %>% t

```

# Session info
```{R}
sessioninfo::session_info()
```