---
title: "AIBL dataset - DMR analysis"
author: "Tiago C. Silva, Lanyu Zhang, Lily Wang"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    highlight: breezedark
    theme: lumen
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
```

# Setup

```{R, message = FALSE, results = 'hide'}
library(dplyr)
library(SummarizedExperiment)
library(GenomicRanges)
library(ExperimentHub)
library(tidyr)
devtools::load_all("~/Documents/packages/coMethDMR/")
cohort <- "AIBL"
dir.base <- "~/TBL Dropbox/Tiago Silva/AD-meta-analysis-blood-samples/"
dir.data <- file.path(dir.base,"datasets/",cohort,"/") 
dir.data.aux <- file.path(dir.data,"/Aux/") 
dir.data.raw <- file.path(dir.data,"/step1_download/") 
dir.data.processed <- file.path(dir.data,"/GSE153712_processed/") 
dir.data.pca <- file.path(dir.data,"/pca_filtering/") 
dir.results.single.cpg <- file.path(dir.base,"analysis_results",cohort,"single_cpg")
dir.results.dmr <- file.path(dir.base,"analysis_results",cohort,"DMR/Logistic_regression_model")
dir.data.dmr <- file.path(dir.data,"/dmr/") 
for(p in grep("dir",ls(),value = T)) dir.create(get(p),recursive = TRUE,showWarnings = FALSE)
```

# Find co-methylated clusters
```{R, eval = FALSE}
devtools::load_all("~/Documents/packages/coMethDMR/")
probes.cluster.all <- coMethDMR::getPredefinedCluster(arrayType = "EPIC",clusterType = "regions")
se <- readRDS(file.path(dir.data.pca, "AIBL_QNBMIQ_PCfiltered_age_at_least_65.RDS"))
```

```{R, eval = FALSE}
coMeth_ls <- CoMethAllRegions(
  dnam = assay(se),      
  betaToM = TRUE,                   
  CpGs_ls = probes.cluster.all,
  arrayType = "EPIC",
  rDropThresh_num = 0.4,
  minPairwiseCorr = NULL,
  method = "spearman",             
  returnAllCpGs = TRUE,              
  output = "all",
  nCores_int = 1,
  progressbar = TRUE
)
save(coMeth_ls,file = file.path(dir.data.dmr,"coMeth_ls.rda"))

```

# Take the median beta-values 

```{R, eval = FALSE}
load(file.path(dir.data.dmr,"coMeth_ls.rda"))
pheno <- data.frame(colData (se))
pheno <- pheno %>% as.data.frame %>% dplyr::select(
  c(disease.status.ch1, sample, Sex, age.pred.Elastic_Net, PLATE,
    B, NK,   CD4T, CD8T,   Mono, Neutro, Eosino
  )
)

pheno$DIAGNOSIS <- factor(
  pheno$disease.status.ch1, 
  levels = c("healthy control", "Mild Cognitive Impairment", "Alzheimer's disease")
)
pheno$DIAGNOSIS_merged <- ifelse(pheno$DIAGNOSIS == "healthy control", 0, 1)
pheno$DIAGNOSIS_merged <- factor(pheno$DIAGNOSIS_merged, levels = c(0, 1))

### Take median of probes in each cluster for each sample
filename <-  file.path(dir.data.dmr,"medianBeta_and_pheno.rda")
beta_mat <- assay(se) %>% as.matrix()
beta_mat <- beta_mat[rownames(beta_mat) %in% unlist(coMeth_ls$coMeth_ls),]

if(!file.exists(filename)){
  medianBetaVal.df <- plyr::ldply(
    coMeth_ls$coMeth_ls[!duplicated(names(coMeth_ls$coMeth_ls))],
    function(probes){
      colMedians(beta_mat[as.character(probes),], na.rm = TRUE)
    },
    .progress = "time"
  )
  medianBetaVal.df$.id <- NULL
  colnames(medianBetaVal.df) <- colnames(beta_mat)
  rownames(medianBetaVal.df) <- names(coMeth_ls$coMeth_ls[!duplicated(names(coMeth_ls$coMeth_ls))])
  save(medianBetaVal.df,pheno, file = filename)
} else {
  load(filename)
}
```

# Logistic regression model

```{R, eval = FALSE}
library(SummarizedExperiment)
library(dplyr)
library(tidyr)
library(lmerTest)

results <- plyr::adply(medianBetaVal.df,.margins = 1, function(region.signal){
  region.signal <- region.signal %>% t %>% as.data.frame() 
  colnames(region.signal) <- "median.beta"
  region.signal$barcodes <- rownames(region.signal)
  pheno$barcodes <- rownames(pheno)
  dat <- dplyr::full_join(region.signal, pheno)
  formula <- paste0(
    "DIAGNOSIS ~ ",
    paste(
      "median.beta",
      "age.pred.Elastic_Net",
      "Sex",
      "as.factor(PLATE)",
      "B",
      "NK",
      "CD4T",
      "CD8T",
      "Mono",
      "Neutro",
      sep = " + "
    )
    )
  
  ################### AD vs. CN --------------------------------------------
  longDat.ad.cn <- subset(dat, disease.status.ch1 == "healthy control"|disease.status.ch1 == "Alzheimer's disease")

  f <- glm(
    formula = formula, 
    data = longDat.ad.cn, 
    family = binomial
  )
  lmer_ad_cn <- coef(summary(f)) [grep("beta",rownames(coef(summary(f)) ),value = TRUE),]
  lmer_ad_cn <- lmer_ad_cn %>% t %>% as.data.frame
  lmer_ad_cn$analysis <- "AD vs CN"
  
  ################### MCI vs. CN -----------------------------------------------
  longDat.mci.cn <- subset(dat, disease.status.ch1 == "healthy control"|disease.status.ch1 == "Mild Cognitive Impairment")
  f <- glm(
    formula = formula, 
    data = longDat.mci.cn, 
    family = binomial
  )
  lmer_cmi_cn <- coef(summary(f)) [grep("beta",rownames(coef(summary(f)) ),value = TRUE),] 
  lmer_cmi_cn <- lmer_cmi_cn %>% t %>% as.data.frame
  lmer_cmi_cn$analysis <- "MCI vs CN"
  return(rbind(lmer_ad_cn ,lmer_cmi_cn))  
}, .progress = "time", .parallel = FALSE,.expand = FALSE,.id = "region")
results$region <- rownames(medianBetaVal.df)[results$region]
readr::write_csv(results, file.path(dir.results.dmr,"DMR_results_all.csv"))
```

# Annotate results 

```{R, eval = FALSE}
results <- readr::read_csv(file.path(dir.results.dmr,"DMR_results_all.csv"))
load(file.path(dir.data.dmr,"coMeth_ls.rda"))
regions <- coMeth_ls$coMeth_ls[!duplicated(names(coMeth_ls$coMeth_ls))]
regions.collapsed <- plyr::llply(regions,.fun =  function(x) paste(x,collapse = ";"),.progress = "time")
results$probeInRegion <- regions.collapsed[match(results$region,names(regions.collapsed))] %>% unlist
results$aux <- gsub("EPIC_Gene_3_200.|EPIC_InterGene_3_200.","",results$region)
results <- results %>% separate(col = "aux",into = c("seqnames","start","end"),remove = TRUE)
results$start <- as.numeric(results$start)
results$end <- as.numeric(results$end)
results.mci <- results[results$analysis == "MCI vs CN",]
results.ad <- results[results$analysis == "AD vs CN",]
```

## Create GREAT annotation for comethylated regions
```{R, eval = FALSE}
library(rGREAT)
load(file.path(dir.data.dmr,"coMeth_ls.rda"))
regions.gr <- results$region %>% unique() %>%
  as.data.frame %>% separate(col = ".",into = c("chr","start","end")) %>% 
  makeGRangesFromDataFrame()

regionsToGenes.list <- plyr::alply(seq(1,length(regions.gr),10000),.margins = 1,.fun = function(start){
  end <- (start + 10000 - 1)
  if(end > length(regions.gr)) end <- length(regions.gr)
  job <- submitGreatJob(regions.gr[start:end], species = "hg19")
  Sys.sleep(70)
  data.frame(plotRegionGeneAssociationGraphs(job))
},.progress = "time")
regionsToGenes <- plyr::rbind.fill(regionsToGenes.list)

regionsToGenes$GREAT_annotation <- ifelse(
  regionsToGenes$distTSS > 0,
  paste0(regionsToGenes$gene, " (+", regionsToGenes$distTSS, ")"),
  paste0(regionsToGenes$gene, " (", regionsToGenes$distTSS, ")"))
regionsToGenes <- regionsToGenes[
  ,c("seqnames", "start", "end", "GREAT_annotation")
]
great <- regionsToGenes %>%
  group_by(seqnames, start, end) %>%
  mutate(GREAT_annotation = paste0(GREAT_annotation,collapse = ";")) %>% unique()
save(great, file = file.path(dir.data.aux,"great_AIBL_coMethDMR_EPIC_array_annotation.rda"))
```

## Add annotation to input cpgs
```{R, eval = FALSE}
load(file.path(dir.data.aux,"great_AIBL_coMethDMR_EPIC_array_annotation.rda"))
load(file.path(dir.data,"../Aux/E073_15_coreMarks_segments.rda"))

add_annotation <- function(result){
  
  result$fdr <- p.adjust(result$`Pr(>|z|)`,method = "fdr")
  message("Annotating E073_15_coreMarks_segments")
  result.gr <- result %>% makeGRangesFromDataFrame(
    start.field = "start",
    end.field = "end",
    seqnames.field = "seqnames"
  )
  hits <- findOverlaps(result.gr, ChmmModels.gr) %>% as.data.frame()
  hits$state <- ChmmModels.gr$state[hits$subjectHits]
  hits$region <- result$region[hits$queryHits]
  result$E073_15_coreMarks_segments_state <- hits$state[match(result$region,hits$region)]
  
  message("Annotating GREAT")
  
  result <- dplyr::left_join(result, great,by = c("seqnames","start","end"))
  
  message("Annotating Island")
  
  result$chrom <- result$seqnames
  result <- AnnotateResults(result, arrayType = "EPIC", nCores_int = 10)
  result$chrom <-  result$chr <- NULL
  result <- result[order(result$`Pr(>|z|)`),]
  return(result)
  
}
results.mci.annotated <- results.mci %>% add_annotation
results.ad.annotated <-  results.ad %>% add_annotation


readr::write_csv(
  results.mci.annotated  %>% as.data.frame(),
  file =  file.path(dir.results.dmr,"DMR_glm_MCI_vs_CN_results_annotated.csv")
)

readr::write_csv(
  results.ad.annotated  %>% as.data.frame(),
  file =  file.path(dir.results.dmr,"DMR_glm_AD_vs_CN_results_annotated.csv")
)

results.mci.annotated.sig <- results.mci.annotated %>% filter(`Pr(>|z|)` < 0.05) 
results.ad.annotated.sig <- results.ad.annotated %>% filter(`Pr(>|z|)` < 0.05) 


readr::write_csv(
  results.ad.annotated.sig  %>% as.data.frame(),
  file = file.path(dir.results.dmr, "DMR_glm_AD_vs_CN_results_annotated_pvalue_cut_off_0_05.csv")
)

readr::write_csv(
  results.mci.annotated.sig  %>% as.data.frame(),
  file = file.path(dir.results.dmr, "DMR_glm_MCI_vs_CN_results_annotated_pvalue_cut_off_0_05.csv")
)

```

# Top 10 regions vs single probes

## AD vs CN
```{R}
sinlge.cpg.ad.vs.cn <- readr::read_csv( 
    file.path(dir.results.single.cpg,"Logistic_regression_model/AIBL_glm_diagnosis_vs_beta_and_covaritates_age_at_least_65_AD_vs_CN_results_annotated_with_inflation.csv"),
    col_types = readr::cols()
)

results.ad.annotated.sig <- readr::read_csv(
  file.path(dir.results.dmr, "DMR_glm_AD_vs_CN_results_annotated_pvalue_cut_off_0_05.csv"),
    col_types = readr::cols()
  )
top10 <- results.ad.annotated.sig[1:10,]
```

```{r loop_print, results = 'asis'}
plyr::a_ply(top10,.margins = 1,.fun = function(x){
  cpgs <- unlist(strsplit(x$probeInRegion,";"))
  cat(paste0("\n### Region ",x$region,"\n"))
  print(x  %>% gt::gt())
  print(
    sinlge.cpg.ad.vs.cn[sinlge.cpg.ad.vs.cn$cpg %in% cpgs,] %>% gt::gt() %>%
      gt::tab_header(
        title = paste0(x$.id)
      )
  )
  
})
```

```{R, include = FALSE}
top10DMR_cpgs <- plyr::alply(top10,.margins = 1,.fun = function(x){
  cpgs <- unlist(strsplit(x$probeInRegion,";"))
    sinlge.cpg.ad.vs.cn[sinlge.cpg.ad.vs.cn$cpg %in% cpgs,]
})
names(top10DMR_cpgs) <- gsub(":|\\.|-","_",top10$region)

writexl::write_xlsx(
  x = c(list("top10_DMRs" = top10),top10DMR_cpgs),
  path =   file.path(dir.results.dmr, "top10_DMR_AD_vs_CN_results_annotated_pvalue_cut_off_0_05.xlsx")
)

```

## MCI vs CN
```{R}
sinlge.cpg.mci.vs.cn <- readr::read_csv(
  file.path(dir.results.single.cpg,"Logistic_regression_model/AIBL_glm_diagnosis_vs_beta_and_covaritates_age_at_least_65_MCI_vs_CN_results_annotated_with_inflation.csv"),
  col_types = readr::cols()
)
```

```{R}
results.mci.annotated.sig <- readr::read_csv(
  file.path(dir.results.dmr, "DMR_glm_MCI_vs_CN_results_annotated_pvalue_cut_off_0_05.csv"),
    col_types = readr::cols()
  )
top10 <- results.mci.annotated.sig[1:10,]
```

```{r, results = 'asis'}
plyr::a_ply(top10,.margins = 1,.fun = function(x){
  cpgs <- unlist(strsplit(x$probeInRegion,";"))
  cat(paste0("\n### Region ",x$region,"\n"))
  print(x  %>% gt::gt())
  print(
    sinlge.cpg.mci.vs.cn[sinlge.cpg.mci.vs.cn$cpg %in% cpgs,] %>% gt::gt() %>%
      gt::tab_header(
        title = paste0(x$.id)
      )
  )
})
```

```{R, include = FALSE}
top10DMR_cpgs <- plyr::alply(top10,.margins = 1,.fun = function(x){
  cpgs <- unlist(strsplit(x$probeInRegion,";"))
    sinlge.cpg.mci.vs.cn[sinlge.cpg.ad.vs.cn$cpg %in% cpgs,]
})
names(top10DMR_cpgs) <- gsub(":|\\.|-","_",top10$region)

writexl::write_xlsx(
  x = c(list("top10_DMRs" = top10),top10DMR_cpgs),
  path = file.path(dir.results.dmr, "top10_DMR_MCI_vs_CN_results_annotated_pvalue_cut_off_0_05.xlsx")
)

```

# Calculate inflation factors and Bacon correction for final results

## Auxiliary functions

```{r inflation,message=FALSE,warning=FALSE}
library(dplyr)
library(bacon)
library(GWASTools)
estimation_of_inflation <- function(data){
  ### 1. Compute genomic inflation factor before bacon adjustment
  data$zvalue <- data$Estimate / data$`Std. Error`
  #data$zvalue <- qnorm(pt(data$tvalue, df))
  data$chisq <- (data$zvalue) ^ 2
  # inflation factor - last term is median from chisq distrn with 1 df  
  inflationFactor <- median(data$chisq,na.rm = TRUE) / qchisq(0.5, 1)
  print("lambda")
  print(inflationFactor)
  # genome-wide sig cpgs
  #sig <- ifelse(data$`Pr(>|t|)` < 2.4e-7, 1, 0)
  # table(sig)  # 1 sig
  ### 2. bacon analysis
  bc <- bacon(
    teststatistics = NULL,
    effectsizes =  data$Estimate,
    standarderrors = data$`Std. Error`,
    na.exclude = TRUE
  )
  # inflation factor
  print("lambda.bacon")
  print(inflation(bc))
  
  ### 3. Create final dataset
  data.with.inflation <- data.frame(
    data,
    Estimate.bacon = bacon::es(bc),
    StdErr.bacon = bacon::se(bc),
    pValue.bacon = pval(bc),
    fdr.bacon = p.adjust(pval(bc), method = "fdr"),
    stringsAsFactors = FALSE
  )
  data.with.inflation <- data.with.inflation %>% dplyr::select(-c(zvalue, chisq))
  return(
    list(
      "data.with.inflation" = data.with.inflation,
      "inflationFactor" =   inflationFactor,
      "estimatedInflation" = inflation(bc)
    )
  )
}

plot_inflation <- function(data, main){
  qqPlot(
    data$data.with.inflation$Pr...z.., ci = TRUE, 
    main = paste0(main)
  ) + text(
    x = 0.5, 
    y = -log10(min(data$data.with.inflation$Pr...z..)) - 0.5,
    labels = bquote(lambda == .(format(data$inflationFactor,digits = 4))),
    pos = 4
  ) + text(
    x = 0.5, 
    y = -log10(min(data$data.with.inflation$Pr...z..)) - 1,
    labels = bquote(lambda[bacon] == .(format(data$estimatedInflation,digits = 4))),
    pos = 4
  )
}
```

## Logistic mixed model

```{R}
results.mci.annotated <- readr::read_csv(
  col_types = readr::cols(),
  file =  file.path(dir.results.dmr,"DMR_glm_MCI_vs_CN_results_annotated.csv")
)

results.ad.annotated <- readr::read_csv(
  col_types = readr::cols(),
  file =  file.path(dir.results.dmr,"DMR_glm_AD_vs_CN_results_annotated.csv")
)
```


```{R}
results.mci.annotated.with.inflation <- estimation_of_inflation(data = results.mci.annotated)
results.ad.annotated.with.inflation <- estimation_of_inflation(data = results.ad.annotated)
```

```{R, eval = F}

readr::write_csv(
  x = results.mci.annotated.with.inflation$data.with.inflation,
  file = file.path(dir.results.dmr,"DMR_glm_MCI_vs_CN_results_annotated_with_inflation.csv")
)

readr::write_csv(
  x = results.ad.annotated.with.inflation$data.with.inflation,
  file = file.path(dir.results.dmr,"DMR_glm_AD_vs_CN_results_annotated_with_inflation.csv")
)

```

```{R}
plot_inflation(results.ad.annotated.with.inflation, "AIBL DMR GLMM - AD vs CN")
```

# Session information
```{R}
devtools::session_info()
```
